Bibat: Batteries-include Bayesian Analysis Template

### 2024-02-26 paper/main.qmd
add automl paper stuff

---
title: "Bibat: Batteries-include Bayesian Analysis Template"
abstract: >
  Bayesian statistical workflow offers a powerful way to learn from data,
  but software software projects that implement complex Bayesian workflows
  in practice are unusual, partly due to the difficulty of orchestrating
  Bayesian statistical software. Bibat addresses this challenge by providing
  a full-featured, scalable Bayesian statistical analysis project using
  an interactive template. Bibat is available on the Python Package index,
  documented at <https://bibat.readthedocs.io/> and developed at <https://github.com/teddygroves/bibat/>. This paper explains the motivation for bibat,
  describes intended usage, discusses bibat's design, compares bibat with similar software, highlights several examples of bibat's use in science and provides links to community resources associated with bibat.
number-sections: true
author:
  - name: "Teddy Groves"
    orcid: 0000-0002-7109-3270
    email: tedgro@biosustain.dtu.dk
    affiliations:
      - id: 1
        name: Danish Technical University
        department: The Novo Nordisk Foundation Center for Biosustainability
        city: Kongens Lyngby
        country: Denmark
    attributes:
      corresponding: true
bibliography: bibliography.bib
keywords: [Bayesian workflow, Methodology, Software]
format:
  typst: default
  html:
    embed-resources: true
  latex: default
---

@@

The most novel part of bibat's design is the introduction of an abstraction
called "fitting mode", which allows bibat project to handle different ways
of fitting a model to a dataset appropriately and flexibly.  This is often
necessary as part of a Bayesian workflow: for example, one might perform
MCMC sampling of both the prior and posterior distributions, perform multiple
leave-out-one-fold fits for cross-validation or need to compare MCMC sampling
with an optimisation-based alternative.

Fitting modes in bibat projects take the form of instances of the class
`FittingMode`. Each fitting mode contains a name, a function that fits a
prepared dataset and instructions for how and where to save the results.
For example, the provided prior sampling fitting mode creates a Stan input
dictionary with the `likelihood` data variable set to `0`, performs MCMC
sampling and writes data to the `InferenceData` group `prior`. This design
allows multiple fits to be stored alongside each other naturally, facilitating
many common Bayesian workflow analyses, and can easily be configured to prevent
unnecessary computations.

@@

@@

# Author contributions

# Acknowledgements

>>>>

---
title: "Bibat: Batteries-include Bayesian Analysis Template"
abstract: >
  Bayesian statistical workflow offers a powerful way to learn from data,
  but software software projects that implement complex Bayesian workflows
  in practice are unusual, partly due to the difficulty of orchestrating
  Bayesian statistical software. Bibat addresses this challenge by providing
  a full-featured, scalable Bayesian statistical analysis project using
  an interactive template. Bibat is available on the Python Package index,
  documented at <https://bibat.readthedocs.io/> and developed at <https://github.com/teddygroves/bibat/>. Bibat is free to use under the MIT license. This paper explains the motivation for bibat, describes intended usage, discusses bibat's design, compares bibat with similar software, highlights several examples of bibat's use in science and provides links to community resources associated with bibat.
number-sections: true
author:
  - name: "Teddy Groves"
    orcid: 0000-0002-7109-3270
    email: tedgro@biosustain.dtu.dk
    affiliations:
      - id: 1
        name: Danish Technical University
        department: The Novo Nordisk Foundation Center for Biosustainability
        city: Kongens Lyngby
        country: Denmark
    attributes:
      corresponding: true
bibliography: bibliography.bib
keywords: [Bayesian workflow, Methodology, Software]
format:
  html:
    embed-resources: true
  pdf:
    template: automl-template.tex
    template-partials:
      - title.tex
---

@@

The most novel part of bibat's design is the introduction of an abstraction
called "fitting mode", which allows bibat projects to handle fitting a model
to a dataset in different ways. This is often necessary as part of a Bayesian
workflow: for example, one might perform MCMC sampling of both the prior
and posterior distributions, or perform multiple leave-out-one-fold fits for
cross-validation, or compare MCMC-based posterior inference with an alternative
computation method.

Fitting modes in bibat projects take the form of instances of the class
`FittingMode`. Each fitting mode contains a name, a function that fits a
prepared dataset and instructions for how and where to save the results. For
example, the provided prior sampling fitting mode is called "prior", contains
a function that runs MCMC sampling with the `likelihood` input variable set
to `0`, returning a `CmdStanMCMC` instance, and performs MCMC sampling and writes data to the `InferenceData` group
`prior`. This design allows multiple fits to be stored alongside each other
naturally, facilitating many common Bayesian workflow analyses, and can easily
be configured to prevent unnecessary computations.

@@

# Limitations

Using bibat effectively requires familiarity with Pydantic, pandera, arviz, Stan
and managing a medium-sized Python project. Many statistical analysis projects
do not require using these tools, for example if data preparation or validation
is trivial, if custom statistical models are not required, or if the analysis
can be carried out by a single script. Practitioners who wish to implement such
Bayesian workflows may prefer to simply write their software from scratch rather
than use bibat, using tools like bambi or brms to ensure that the software
challenge remains manageable.

Similarly, bibat includes features that not all analyses will need, such as
accommodating plural inferences, fitting modes and datasets. As a result, bibat
is not optimal for projects that are known in advance to be singular. This could
happen, for example, because there is no uncertainty as to the necessary data
manipulations, or because the aim of the project is to maintain and streamline
an existing statistical model rather than to compare it with others.

While inclusion of potentially unneeded features is an important limitation
of bibat, it is worth noting that it is often difficult to predict in advance
what features a Bayesian workflow project will require, and that it can
be time-consuming to re-write a project after belatedly realising that an
unsupported feature is needed. We therefore think there is value in using bibat
as a default.

Another limitation of bibat is that it makes many opinionated choices about
which tools to use. In particular, languages other than Python, inference
frameworks other than Stan and validation frameworks other than Pydantic are
not supported. We think this limitation is best addressed by the development of
additional analysis templates.

@@

# Broader impact statement

After careful reflection, the authors have determined that this work presents no
notable negative impacts to society or the environment.

\begin{acknowledgements}

This research was funded by the Novo Nordisk Foundation (Grant numbers
NNF20CC0035580 NNF14OC0009473).

\end{acknowledgements}
